Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-03-21 paper/ms.tex
Editing summary section

The asteroseismic oscillations of thousands of stars were measured using light
curves from the \kepler\ mission \citep{Gilliland:2010, Chaplin:2011,
Chaplin:2013, Stello:2013} and asteroseismology is a key science driver for
many of the upcoming large scale photometric surveys \citep{Campante:2016,
Rauer:2014, Gould:2015}.
Most asteroseismic analyses have been limited to relatively high
signal-to-noise oscillations because the standard methods based on statistics
of the empirical periodogram of the data cannot be used to formally propagate
the measurement uncertainties to the constraints on physical parameters
(\dfmtodo{CITE}) but more sophisticated methods that compute the likelihood
function in the time domain scale poorly to state-of-the-art datasets
\citep{Brewer:2009}.

@@

Although we have in mind applications of this fast method to model stellar
variability, the method is general for one-dimensional GP problems, and may be
applied to other problems.
Accreting black holes show time series which may be modeled by correlated
noise \citep{Kelly:2014};  indeed, this was the motivation for the original
technique developed by Rybicki \& Press \citep{Rybicki:1992,Rybicki:1995}.
This approach may be broadly used for characterizing quasar variability
\citep{MacLeod:2010}, measuring time lags with reverberation mapping
\citep{Zu:2011,Pancoast:2014}, and modeling time delays in multiply-imaged
gravitationally-lensed systems \citep{Press:1998}.

Outside of astronomy, this technique may have application to seismology
\citep{Robinson:1967},

There are three primary applications we envision for the \celerite\ formalism:
1).\ modeling the variability of an astrophysical system to infer it
properties; 2).\ accounting for astrophysical variability as a source of noise
when trying to detect additional phenomena; 3).\ interpolating or
extrapolating variability to future or missing times.
Some examples of the first are determining the characteristic variability
timescale of a quasar \citep{Kelly:2014}, measuring the asteroseismic
variation of a star \citep{Brewer:2009,Corsaro:2014}, measurement of
granulation noise \citep{Bastien:2013,Kallinger:2016}, detection of
quasi-periodic variability in a high-energy source \citep{McAllister:2016}, or
classification of variable objects \citep{Zinn:2016}.
Examples of the second are detecting and characterizing transiting exoplanets
\citep[for example,][]{Gibson:2012, Barclay:2015, Evans:2015, Aigrain:2016,
Grunblatt:2016, Luger:2016}
and correcting for stellar activity in radial velocity measurements
\citep[for example,][]{Haywood:2014, Rajpaul:2015}.
Examples of the third are measuring time delays in multiply-imaged
gravitationally lensed sources \citep{Tewes:2013} and reverberation mapping of
AGN.

Our background is in studying transiting exoplanets, a field which has only
recently begun to adopt full covariance matrices in analyzing the noise in
transiting planet lightcurves.
One promising early attempt at accounting for correlated noise was the wavelet
approach of \citet{Carter:2009}.
Their technique allowed for a power-law spectrum for the noise scaling as
$\omega^{-\alpha}$, which unfortunately is not a normalizable, nor
physically-accurate, description of stellar variability.
Their approach runs in ${\cal O}(N)$ time for $\alpha=1$, which is referred to
as `flicker', $1/r$, or pink noise;  however, this particular type of noise
overpredicts the power spectrum at low frequency compared to the Solar
granulation power spectrum.
It also cannot describe quasi-periodic noise.

@@

In the future we hope to expand the \celerite\ technique to multi-dimensional
datasets.  The formalism presented could, for example, be used to model
the covariance between multiple measured quantities, such as in regression of photometric
time series versus astrometric precision when correcting for systematic errors
due to detector sensitivity variations \citep{Aigrain:2016}, in correcting
radial velocity datasets for stellar activity \citep{Haywood:2014,Rajpaul:2015},
or in detection of gravitational waves in the presence of systematic errors
\citep{Moore:2016}.

>>>>

The asteroseismic oscillations of thousands of stars were measured using light
curves from the \kepler\ mission \citep{Gilliland:2010, Chaplin:2011,
Chaplin:2013, Stello:2013} and asteroseismology is a key science driver for
many of the upcoming large scale photometric surveys \citep{Campante:2016,
Rauer:2014, Gould:2015}.
Most asteroseismic analyses have been limited to relatively high
signal-to-noise oscillations because the standard methods based on statistics
of the empirical periodogram of the data cannot be used to formally propagate
the measurement uncertainties to the constraints on physical parameters
(\dfmtodo{CITE}) but more sophisticated methods that compute the likelihood
function in the time domain scale poorly to state-of-the-art datasets
\citep{Brewer:2009,Corsaro:2014}.

@@

The computational burden of Gaussian process regression has limited its application
to small astronomical time series.  With the linear scaling of \celerite\, we
envision the application of Gaussian processes may be expanded to much larger 
datasets.  Despite the restrictive form of the \celerite\ kernel, with a 
sufficient number of components it is flexible enough to describe a wide 
range of astrophysical variability.  In fact, the relation of the \celerite\ kernel 
to the damped, driven harmonic oscillator matches simple models of astrophysical 
variability, and makes the parameterization interpretable in terms of resonant 
frequency, amplitude, and quality factor.  Perhaps the most significant drawback 
is the $J^2$ scaling of the solver;  however, for many problems a small value of $J$ 
suffices.

We envision two general varieties of applications:  those interested in the
the \celerite\ kernel parameters (so-called ``hyper-parameters") which describe the
astrophysical variability, versus those interested in characterizing the mean 
model, treating the correlated noise as a nuisance.  Some users may
be interested in both applications, while other problems may require interpolation
or extrapolation beyond the dataset that is being modeled.

%Some examples of the first are determining the characteristic variability
%timescale of a quasar \citep{Kelly:2014}, measuring the asteroseismic
%variation of a star \citep{Brewer:2009,Corsaro:2014}, measurement of
%granulation noise \citep{Bastien:2013,Kallinger:2016}, detection of
%quasi-periodic variability in a high-energy source \citep{McAllister:2016}, or
%classification of variable objects \citep{Zinn:2016}.
%Examples of the second are detecting and characterizing transiting exoplanets
%\citep[for example,][]{Gibson:2012, Barclay:2015, Evans:2015, Aigrain:2016,
%Grunblatt:2016, Luger:2016}
%and correcting for stellar activity in radial velocity measurements
%\citep[for example,][]{Haywood:2014, Rajpaul:2015}.
%Examples of the third are measuring time delays in multiply-imaged
%gravitationally lensed sources \citep{Tewes:2013} and reverberation mapping of
%AGN.

Our background is in studying transiting exoplanets, a field which has only
recently begun to adopt full covariance matrices in analyzing the noise in
stellar lightcurves when detecting or characterizing transiting planets
\citep[for example,][]{Gibson:2012, Barclay:2015, Evans:2015, Aigrain:2016, 
Grunblatt:2016, Luger:2016}.
One promising early attempt at accounting for correlated noise was the wavelet
approach of \citet{Carter:2009}.
Their technique allowed for a power-law spectrum for the noise scaling as
$\omega^{-\alpha}$, which unfortunately is not a normalizable, nor
physically-accurate, description of stellar variability.
Their approach runs in ${\cal O}(N)$ time for $\alpha=1$, which is referred to
as `flicker', $1/r$, or pink noise;  however, this particular type of noise
overpredicts the power spectrum at low frequency compared to the Solar
granulation power spectrum.  It also cannot describe quasi-periodic noise.

@@

Although we have in mind applications of this fast method to model stellar
variability, the method is general for one-dimensional GP problems, and may be
applied to other problems.
Accreting black holes show time series which may be modeled by correlated
noise \citep{Kelly:2014};  indeed, this was the motivation for the original
technique developed by Rybicki \& Press \citep{Rybicki:1992,Rybicki:1995}.
This approach may be broadly used for characterizing quasar variability
\citep{MacLeod:2010}, measuring time lags with reverberation mapping
\citep{Zu:2011,Pancoast:2014}, modeling time delays in multiply-imaged
gravitationally-lensed systems \citep{Press:1998}, characterizing quasi-periodic 
variability in a high-energy source \citep{McAllister:2016}, or
classification of variable objects \citep{Zinn:2016}.
Outside of astronomy, this technique may have application to terrestrial 
seismology \citep{Robinson:1967}.
The \celerite\ kernel may be used in lieu of a Gaussian power-spectrum for
non-parametric data modeling \citep{Wilson:2013}.

In the future we hope to expand the \celerite\ technique to multi-dimensional
datasets.  The formalism presented could, for example, be used to model
the covariance between multiple measured quantities, such as in regression of photometric
time series versus astrometric precision when correcting for systematic errors
due to detector sensitivity variations \citep{Aigrain:2016}, in correcting
radial velocity datasets for stellar activity \citep{Haywood:2014,Rajpaul:2015},
or in detection of gravitational waves in the presence of systematic errors
\citep{Moore:2016}.  Other extensions that may be useful are fast simulation
of \celerite\ processes, prediction and/or interpolation of datasets, and
extended the \celerite\ kernel to time-varying kernels.
